Layered Grammar for Alluvial Plots

### 2020-01-18 paper.rmd
revise concepts section of joss paper

Visualizations of flow processes have long encoded magnitudes as ribbon widths, constituting a type called Sankey diagrams [@Schmidt2008].
A widely-used subtype for longitudinal categorical data represent categories as nodes threaded by edges that represent the trajectories and magnitudes of cases [@Riehmann2005].
Their design anticipated parallel sets plots, which were adapted from parallel coordinates plots [@Inselberg1987;@Wegman1990] to visualize multivariate categorical data, and which represent cohorts of equivalent cases as ribbons connecting categories represented as boxes [@Kosara2006].
These in turn anticipated "alluvial diagrams", proposed to visualize changes in case memberships across successive cross-sections [@Rosvall2010].
Several R packages have been developed to generate these diagrams, in particular alluvial [@Bojanowski2016], ggparallel [@Hofmann2013], ggforce [@Pedersen2019], ggalluvial [@Brunson2019], and ggpcp [@Ge2019].

Sankey, parallel sets, and alluvial diagrams are often conflated, and there is currently no consensus on what features are distinctive to each type.
Moreover, their graphical elements go by a variety of names, often interchangeably.
In order to more clearly describe the features of ggalluvial in relation to similar packages, it is useful to adopt a formal demarcation among these diagram types.

_Statistical graphics_ (here also simply called "plots") are diagrams that communicate statistical information using graphical methods [@Friendly2005] and, more narrowly, are uniquely determined from data by a fixed set of plotting rules [@Wilkinson2006].
By design, graphics produced by ggplot2 extensions (and by the alluvial package) are plots: The stat, geom, and other layers of a ggplot object exactly reproduce a graphic from data (under the same parameter settings).[^exceptions]
Sankey diagrams are much more flexible: The earliest engine efficiency diagrams in this tradition could take a variety of forms to depict the same energy flow and were differently annotated for different audiences [@Schmidt2008], and software implementations may be interactive and enable manual adjustments of graphical elements [@Riehmann2005].
Moreover, the plots generated by these extensions are suitable for a wide variety of data types, whereas Sankey diagrams are still overwhelmingly used to represent flow. Arguably, the extensions are better understood as belonging to a different type of diagram.

@@

Parallel sets plots might be viewed as a special subtype of Sankey diagram. In turn, the plots produced by the aforecited packages are all parallel sets plots: Cohorts are partitioned into categories at each axis and connected by ribbons whose widths encode their magnitudes.[^hammock]
They vary in the shapes of ribbons, the arrangements of boxes, and the presence of gaps between boxes at the same axis.
The geoms of ggalluvial offer only a one-dimensional continuum between straight lines and x-splines for the ribbon shapes, though alternatives are planned.
The stats vertically arrange the elements (boxes and ribbons) at each axis. ggalluvial provides more levers of control over these statistical transformations, thereby over the messages conveyed by the plot, than the other packages.[^easy]
Notably, in ggalluvial the plot elements are rendered by separate layers, following the additive (`+`) syntax of ggplot2, a feature shared by ggforce and ggpcp but not by ggparallel.

@@

The ggalluvial package adopts the term _alluvial plot_ for the subtype of parallel sets plots it produces,[^plots] which are distinguished by two features: a strict order on the stacked elements at each axis, including both the values of the discrete variables and the ribbons connecting cases or cohorts between them; and a meaningful positional dimension along which these elements are stacked, which precludes inserting gaps between them. The first feature is shared by the other packages but are not essential to parallel sets plots; such plots could arrange boxes corresponding to repeated categorical decompositions differently at different axes. While most of the packages separate boxes at each axis with gaps, these can be reduced to zero, so that each package can create alluvial plots. (ggparallel and ggalluvial alone _only_ produce alluvial plots.)
This feature is particularly important to some applications and, in my view, can fundamentally change the way a plot is interpreted.
It is for this reason that i believe the new geological terminology is warranted.

@@

While most uses might be served equally well by other parallel sets plots or Sankey diagrams, ggalluvial seems exceptionally well-suited to three settings: repeated ordinal measures data, incomplete longitudinal data, and signed categorical data.[^opinion]

>>>>

Visualizations of flow processes have long encoded magnitudes as ribbon widths, constituting a type called Sankey diagrams [@Schmidt2008].
A widely-used subtype for longitudinal categorical data represent categories as nodes threaded by edges that represent the trajectories and magnitudes of cases [@Riehmann2005].
Their design anticipated parallel sets plots, which were adapted from parallel coordinates plots [@Inselberg1987;@Wegman1990] to visualize multivariate categorical data, and which represent cohorts of equivalent cases as ribbons connecting categories represented as boxes [@Kosara2006].
These in turn anticipated "alluvial diagrams", proposed to visualize changes in case memberships across successive cross-sections [@Rosvall2010].
Several R packages have been developed to generate diagrams of these types, including riverplot [@Weiner2017], networkD3 [@Allaire2017], sankey [@Csardi2017], alluvial [@Bojanowski2016], ggparallel [@Hofmann2013], ggforce [@Pedersen2019], ggalluvial [@Brunson2019], and ggpcp [@Ge2019].

Sankey, parallel sets, and alluvial diagrams are often conflated, and there is currently no consensus on what features are distinctive to each type.
Moreover, their graphical elements go by a variety of names, often interchangeably.
In order to more clearly describe the features of ggalluvial in relation to similar packages, it is useful to adopt a careful demarcation among these diagram types.

_Statistical graphics_ (here also simply called "plots") are diagrams that communicate statistical information using graphical methods [@Friendly2005] and, more narrowly, are uniquely determined from data by a fixed set of plotting rules [@Wilkinson2006].
By design, graphics produced by ggplot2 extensions are plots: The stat, geom, and other layers of a ggplot object exactly reproduce a graphic from data (under the same parameter settings).[^exceptions]
Sankey diagrams are much more flexible.
The earliest engine efficiency diagrams in this tradition could take a variety of forms to depict the same energy flow and were differently annotated for different audiences [@Schmidt2008].
Software implementations may use heuristic algorithms to position graphical elements [@Allaire2017;@Csardi2017] or enable users to manually, even interactively, adjust them [@Riehmann2005;@Allaire2017;@Weiner2017].
Paradoxically, Sankey diagrams are overwhelmingly used to represent flow, whereas the aforecited ggplot2 extensions are used to visualize a wide variety of data types.
Arguably, the extensions are better understood as producing a different type of diagram.

@@

Parallel sets plots might be viewed as a special subtype of Sankey diagram with the following features: Ribbons proceed monotonically along one dimension, and e very ribbon encounters a box at every axis. These graphical constraints correspond to combinatorial constraints on the data, which amount to a id--key--value structure in which every id--key pair takes exactly one value (possibly 0 or `NA`).
In this sense, the plots produced by the ggplot2 extensions (and by the alluvial package) are parallel sets plots: Cohorts are partitioned into categories at each axis and connected by ribbons whose widths encode their magnitudes.[^hammock]

The plots produced still vary---in the shapes of ribbons, the arrangements of boxes, and the presence of gaps between boxes at the same axis.
The exceptional geoms of ggparallel offer common-angle as well as linear ribbons. Those of alluvial, ggforce, ggalluvial, and ggpcp offer a one-dimensional continuum between straight and x-spline ribbons.
The stats vertically arrange the elements (boxes and ribbons) at each axis.
These distinct elements are rendered by separate layers in ggforce, ggalluvial, and ggpcp, following the additive (`+`) syntax of ggplot2.
ggalluvial provides more levers of control over these statistical transformations, thereby over the messages conveyed by the plot, than the other packages.[^easy]

@@

The ggalluvial package adopts the term _alluvial plot_ for the subtype of parallel sets plots it produces,[^plots] with the geological terminology itemized above.
These alluvial plots are distinguished by two features: a strict order on the stacked elements at each axis, including both the values of the discrete variables and the ribbons connecting cases or cohorts between them; and a meaningful positional dimension along which these elements are stacked, which precludes inserting gaps between them. The first feature is shared by the other packages but is not essential to parallel sets plots; such plots could arrange boxes corresponding to repeated categorical decompositions differently at different axes. While most of the packages separate boxes at each axis with gaps, these can be reduced to zero, so that each package can create alluvial plots. (ggparallel and ggalluvial alone _only_ produce alluvial plots.)
This feature is particularly important to some applications and, in my view, can fundamentally change the way a plot is interpreted.
It is for this reason that i believe the typology and terminology are warranted.

@@

While most uses might be served equally well by other parallel sets plots or Sankey diagrams, alluvial plots seem exceptionally well-suited to three settings: repeated ordinal measures data, incomplete longitudinal data, and signed categorical data.[^opinion]
